---
title: "Tariff Exposure Teaching Series"
subtitle: "Session 2 - Measurement Reconstruction, NAICS Audit, and Mapping"
author: "Ian Helfrich"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
    number_sections: false
    df_print: paged
    css: workbook_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

<div class="session-banner">
<b>Session focus:</b> Verify that the exposure pipeline is internally coherent, catch NAICS problems early, and use maps as diagnostics.
</div>

## 1) Environment and Packages

```{r env-check}
if (file.exists("retaliatory-tariff.Rproj")) {
  project_root <- "."
} else if (file.exists("../retaliatory-tariff.Rproj")) {
  project_root <- ".."
} else {
  stop("Could not find retaliatory-tariff.Rproj. Open this file inside the project folder.")
}

setwd(project_root)
knitr::opts_knit$set(root.dir = normalizePath(getwd()))
cat("Working directory:", normalizePath(getwd()), "\n")
```

```{r libraries}
needed <- c(
  "readr", "dplyr", "tidyr", "stringr", "ggplot2", "scales", "knitr", "sf", "tigris"
)
missing_pkgs <- needed[!vapply(needed, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  stop("Install missing packages first:\n", paste0("install.packages('", missing_pkgs, "')", collapse = "\n"))
}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(knitr)
})
```

## 2) Quick Formula Refresher (with Color Anchors)

<div class="math-legend">
  <span class="pill c-share">Blue = local share</span>
  <span class="pill c-shock">Orange = national tariff shock</span>
  <span class="pill c-outcome">Green = county exposure output</span>
</div>

\[
\color{DarkGreen}{\text{Exposure}_{ct}}
=
\sum_i
\underbrace{\color{MidnightBlue}{\left(\frac{L_{cit}}{L_{ct}}\right)}}_{\color{MidnightBlue}{\text{share}}}
\times
\underbrace{\color{Sienna}{\left(X^{US}_{it} \times \tau_{it}\right)}}_{\color{Sienna}{\text{shock}}}
\]

## 3) Reconstruct Exposure From Components

### Framing {.wb-framing}

If we cannot rebuild the county-year exposure from industry-level components, we should not trust any downstream model.

```{r load-main-data}
panel <- read_csv(
  "data/public/derived/county_tariff_exposure_panel_with_votes_2017_2023.csv",
  show_col_types = FALSE
) %>%
  mutate(
    county_fips = str_pad(as.character(county_fips), 5, pad = "0"),
    year = as.integer(year)
  )

components <- read_csv(
  "data/public/derived/county_industry_tariff_panel_2017_2023.csv",
  show_col_types = FALSE
) %>%
  mutate(
    county_fips = str_pad(as.character(county_fips), 5, pad = "0"),
    year = as.integer(year)
  )
```

```{r hoonah-rebuild}
focus <- components %>%
  filter(str_detect(county_name, "Hoonah-Angoon")) %>%
  distinct(county_fips, county_name) %>%
  slice(1)

focus_fips <- focus$county_fips[[1]]
focus_name <- focus$county_name[[1]]
focus_year <- max(panel$year, na.rm = TRUE)

cat("County:", focus_name, "| FIPS:", focus_fips, "| Year:", focus_year, "\n")

focus_tbl <- components %>%
  filter(county_fips == focus_fips, year == focus_year) %>%
  mutate(reconstructed_component = emp_share * export_value_usd * tariff_rate_mean) %>%
  select(
    naics2c, emp_share, export_value_usd, tariff_rate_mean,
    county_tariff_component_mean, reconstructed_component
  ) %>%
  arrange(desc(reconstructed_component))

kable(slice_head(focus_tbl, n = 12), digits = 5)
```

```{r consistency-check}
reconstructed <- components %>%
  group_by(county_fips, year) %>%
  summarise(reconstructed_exposure = sum(county_tariff_component_mean, na.rm = TRUE), .groups = "drop")

consistency <- panel %>%
  select(county_fips, year, county_tariff_exposure_mean) %>%
  inner_join(reconstructed, by = c("county_fips", "year")) %>%
  mutate(abs_diff = abs(county_tariff_exposure_mean - reconstructed_exposure))

consistency %>%
  summarise(
    median_abs_diff = median(abs_diff, na.rm = TRUE),
    max_abs_diff = max(abs_diff, na.rm = TRUE)
  ) %>%
  kable(digits = 6)
```

### Checkpoint {.wb-checkpoint}

What does a very small `max_abs_diff` tell us about pipeline reliability?

## 4) NAICS Audit Before Modeling

```{r naics-audit}
emp_panel <- read_csv(
  "data/public/derived/county_industry_emp_share_2017_2023.csv",
  show_col_types = FALSE
) %>%
  mutate(
    county_fips = str_pad(as.character(county_fips), 5, pad = "0"),
    naics2c = as.character(naics2c),
    valid_naics = str_detect(naics2c, "^(00|[0-9]{2}|31-33|44-45|48-49|99)$"),
    emp = as.numeric(emp),
    emp_share = as.numeric(emp_share)
  )

naics_issues <- emp_panel %>%
  mutate(
    issue = case_when(
      !valid_naics ~ "bad_naics_format",
      is.na(emp) ~ "missing_emp",
      is.na(emp_share) ~ "missing_emp_share",
      !is.finite(emp_share) ~ "nonfinite_emp_share",
      TRUE ~ "ok"
    )
  ) %>%
  filter(issue != "ok") %>%
  count(year, issue, naics2c, sort = TRUE, name = "rows")

kable(slice_head(naics_issues, n = 25))
```

```{r write-naics-audit}
dir.create("output/analysis/tables", recursive = TRUE, showWarnings = FALSE)
write_csv(naics_issues, "output/analysis/tables/naics_problem_entries_session2.csv")
cat("Wrote output/analysis/tables/naics_problem_entries_session2.csv\n")
```

## 5) Map Diagnostic (Static, Readable)

### Framing {.wb-framing}

A map should answer: where are high values, how smooth or clustered are they, and where are missing values?

```{r map-exposure, fig.height=7, fig.width=12}
latest_year <- max(panel$year, na.rm = TRUE)
latest <- panel %>% filter(year == latest_year)

options(tigris_use_cache = TRUE)

counties_sf <- tigris::counties(cb = TRUE, year = 2023, class = "sf") %>%
  sf::st_as_sf() %>%
  select(GEOID, geometry)

map_df <- counties_sf %>%
  left_join(
    latest %>%
      transmute(GEOID = county_fips, exposure_per_worker = county_tariff_exposure_per_worker),
    by = "GEOID"
  ) %>%
  filter(!str_sub(GEOID, 1, 2) %in% c("02", "15", "60", "66", "69", "72", "78"))

ggplot(map_df) +
  geom_sf(aes(fill = exposure_per_worker), color = NA) +
  scale_fill_viridis_c(option = "C", trans = "log1p", na.value = "grey92") +
  labs(
    title = paste0("County tariff exposure per worker (", latest_year, ")"),
    subtitle = "Log-scaled fill to keep the full distribution visible",
    fill = "Exposure / worker"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())
```

### Teaching takeaway {.wb-takeaway}

This map is a diagnostic layer, not the final claim. It tells us what patterns need explanation, not which explanation is automatically true.

## 6) Session Wrap

### Checkpoint {.wb-checkpoint}

Before Session 3, be ready to explain:

- How we validated the formula by reconstruction
- Which NAICS issues appeared and why they matter
- What the map revealed and what it did not prove

<p class="small-note">Next session: regression ladder, uncertainty discipline, and optional campaign-finance controls.</p>
