---
title: "Tariff Exposure Teaching Series"
subtitle: "Session 1 - Foundations, Intuition, and Data Structure"
author: "Ian Helfrich"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
    number_sections: false
    df_print: paged
    css: workbook_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

<div class="session-banner">
<b>Session focus:</b> Build the core mental model for tariff exposure, make the formula intuitive, and connect each R step to a research decision.
</div>

## 1) How To Run This Session

### Framing {.wb-framing}

This session is not about speed. It is about learning to think clearly about what each object means before we model anything.

A good pace is:

- Read the short framing paragraph out loud.
- Run the code chunk.
- Ask one checkpoint question before moving on.

If you hit confusion, pause there. That is where learning happens.

### Teaching takeaway {.wb-takeaway}

The goal is not memorizing code. The goal is learning how to map an economic idea into a defensible data object.

## 2) The Core Question And Why It Is A Good One

### Framing {.wb-framing}

Our question is:

Did retaliatory tariffs hit politically salient places more than others?

Why this is strong:

- The policy shock is real and externally observed.
- The geography is local and measurable.
- The political outcomes are public and county-indexed.

That gives us a clean empirical spine: shock, place, outcome.

### Checkpoint {.wb-checkpoint}

In plain language, what would count as evidence for this question, and what would count as evidence against it?

## 3) Vocabulary You Need Before Touching The Formula

### Framing {.wb-framing}

A lot of confusion comes from vocabulary drift. Pin the terms down first.

- <span class="c-shock">Shock</span>: a policy or market event that changes exposure conditions from outside the county.
- <span class="c-share">Share</span>: how concentrated local employment is in a given industry.
- <span class="c-outcome">Exposure</span>: predicted local vulnerability implied by shock x share.
- `county_fips`: a 5-digit county ID (state + county code).
- `NAICS`: industry classification labels.

Once these are stable, the code becomes much easier to read.

## 4) The Formula, Underbraces, And Intuition

<div class="math-legend">
  <span class="pill c-share">Blue = local share term</span>
  <span class="pill c-shock">Orange = national shock term</span>
  <span class="pill c-outcome">Green = county exposure output</span>
</div>

\[
\color{DarkGreen}{\text{Exposure}_{ct}}
=
\sum_i
\underbrace{\color{MidnightBlue}{\left(\frac{L_{cit}}{L_{ct}}\right)}}_{\color{MidnightBlue}{\text{local industry share}}}
\times
\underbrace{\color{Sienna}{\left(X^{US}_{it} \times \tau_{it}\right)}}_{\color{Sienna}{\text{national tariff-exposed shock}}}
\]

Read this as a sentence:

County exposure equals the sum, over industries, of local workforce reliance times national tariff pressure in that same industry.

What each piece is doing:

- <span class="c-share">Blue share term</span>: asks, "How much does this county rely on industry \(i\)?"
- <span class="c-shock">Orange shock term</span>: asks, "How hard is industry \(i\) being hit nationally by tariffs?"
- <span class="c-outcome">Green output</span>: aggregates those interactions into one county-year score.

### Teaching takeaway {.wb-takeaway}

The formula is not mystical. It is a weighted average idea where the weights come from local employment structure.

### Checkpoint {.wb-checkpoint}

If two counties face the same national shock vector, what makes their exposure scores different?

## 5) One Tiny Numerical Example (So The Formula Feels Real)

### Framing {.wb-framing}

Before real data, build trust with a tiny hand-checkable example.

```{r toy-county-example}
library(dplyr)
library(tibble)
library(knitr)

toy_county <- tibble(
  industry = c("Soy", "Machinery", "Chemicals"),
  local_emp_share = c(0.20, 0.35, 0.45),
  national_exports_usd = c(8e9, 14e9, 10e9),
  tariff_rate = c(0.25, 0.10, 0.05)
) %>%
  mutate(
    national_shock = national_exports_usd * tariff_rate,
    contribution = local_emp_share * national_shock
  )

toy_county %>%
  mutate(across(c(local_emp_share, tariff_rate), round, 3)) %>%
  kable()
```

```{r toy-county-total}
toy_county %>%
  summarise(
    exposure_total = sum(contribution),
    share_sum = sum(local_emp_share)
  ) %>%
  kable(digits = 2)
```

Interpretation:

- `share_sum` should be near 1 in a clean complete setup.
- `exposure_total` is the county score implied by its industrial mix and national tariff pressure.

This is exactly the same logic as the full pipeline, just miniaturized.

### Checkpoint {.wb-checkpoint}

Which industry contributes the most in this toy table, and is it because of local share, national shock size, or both?

## 6) Two Intuition Drills That Prevent Bad Inference

### 6.1 The Laredo Problem (Geography)

### Framing {.wb-framing}

Production location and border-exit location can differ sharply. If we treat them as the same, we misallocate exposure.

```{r laredo-toy}
toy_geo <- tibble(
  county = c("Detroit area", "Laredo area"),
  manufacturing_workers = c(180000, 12000),
  border_exit_value_usd = c(2000000000, 12000000000)
)

toy_geo %>% kable()
```

How to teach this table:

- Laredo can have very high exit value because goods cross there.
- Detroit can have much higher production-side labor exposure because goods are produced there.
- Exit nodes and production nodes are different economic objects.

### 6.2 The Los Angeles Problem (Scale)

### Framing {.wb-framing}

Total dollar exposure can overstate impact in very large labor markets and understate intensity in smaller ones.

\[
\underbrace{\color{DarkGreen}{\text{Exposure per worker}_{ct}}}_{\color{DarkGreen}{\text{intensity measure}}}
=
\frac{
\underbrace{\color{MidnightBlue}{\text{Exposure}_{ct}}}_{\color{MidnightBlue}{\text{total burden}}}
}{
\underbrace{\color{Sienna}{L_{ct}}}_{\color{Sienna}{\text{county employment base}}}
}
\]

```{r scale-toy}
scale_toy <- tibble(
  county = c("Small county", "Large county"),
  total_exposure = c(5e7, 5e7),
  county_employment = c(25000, 900000)
) %>%
  mutate(exposure_per_worker = total_exposure / county_employment)

scale_toy %>% kable(digits = 2)
```

Same total exposure, very different per-worker intensity.

### Checkpoint {.wb-checkpoint}

When writing results, how would you explain a county that is high in levels but not high in per-worker terms?

## 7) R Setup: Why We Check Paths And Packages First

### Framing {.wb-framing}

Most reproducibility failures are environment failures, not statistical failures.

```{r env-check}
if (file.exists("retaliatory-tariff.Rproj")) {
  project_root <- "."
} else if (file.exists("../retaliatory-tariff.Rproj")) {
  project_root <- ".."
} else {
  stop("Could not find retaliatory-tariff.Rproj. Open this file inside the project folder.")
}

setwd(project_root)
knitr::opts_knit$set(root.dir = normalizePath(getwd()))
cat("Working directory:", normalizePath(getwd()), "\n")
```

How to explain this chunk live:

- We look for the project file first so relative paths are stable.
- `setwd(project_root)` ensures subsequent file reads point to the same place for anyone running this.
- `opts_knit$set(root.dir=...)` keeps knitting behavior consistent.

```{r libraries}
needed <- c("readr", "dplyr", "tidyr", "stringr", "ggplot2", "scales", "knitr")
missing_pkgs <- needed[!vapply(needed, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  stop("Install missing packages first:\n", paste0("install.packages('", missing_pkgs, "')", collapse = "\n"))
}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(knitr)
})
```

How to explain this chunk live:

- `needed` declares dependencies explicitly.
- `requireNamespace(..., quietly = TRUE)` checks availability without loading everything.
- `stop(...)` fails early with a useful message instead of failing later in a confusing way.

### Teaching takeaway {.wb-takeaway}

Failing early is a kindness to future-you.

## 8) Load The County Panel And Read It Like A Researcher

```{r load-panel}
panel <- read_csv(
  "data/public/derived/county_tariff_exposure_panel_with_votes_2017_2023.csv",
  show_col_types = FALSE
) %>%
  mutate(
    county_fips = stringr::str_pad(as.character(county_fips), 5, pad = "0"),
    year = as.integer(year)
  )

panel %>%
  summarise(
    rows = n(),
    years = n_distinct(year),
    counties = n_distinct(county_fips),
    min_year = min(year, na.rm = TRUE),
    max_year = max(year, na.rm = TRUE)
  ) %>%
  kable()
```

What this is doing technically:

- `read_csv(...)` imports the panel object.
- `str_pad(...)` standardizes FIPS codes to 5 digits, preserving leading zeros.
- `as.integer(year)` makes year explicit numeric time index.
- `summarise(...)` gives a quick integrity read: row count, temporal span, and cross-sectional coverage.

```{r panel-coverage}
panel %>%
  count(year, name = "county_rows") %>%
  kable()
```

How to discuss this table:

- We check whether coverage is stable across years.
- Sharp drops may indicate pipeline gaps, not economic effects.
- Data coverage diagnostics are part of causal discipline.

### Checkpoint {.wb-checkpoint}

Why can unstable county coverage across years distort comparisons, even before any regression?

## 9) Level Vs Intensity In The Real Panel

```{r level-vs-intensity-quick}
latest_year <- max(panel$year, na.rm = TRUE)
latest <- panel %>% filter(year == latest_year)

level_top <- latest %>%
  arrange(desc(county_tariff_exposure_mean)) %>%
  select(county_name, county_fips, county_tariff_exposure_mean, county_tariff_exposure_per_worker) %>%
  slice_head(n = 10)

intensity_top <- latest %>%
  arrange(desc(county_tariff_exposure_per_worker)) %>%
  select(county_name, county_fips, county_tariff_exposure_mean, county_tariff_exposure_per_worker) %>%
  slice_head(n = 10)

kable(level_top, digits = 3, caption = "Top 10 by exposure level")
kable(intensity_top, digits = 3, caption = "Top 10 by exposure per worker")
```

```{r overlap-check}
overlap_n <- dplyr::inner_join(
  level_top %>% select(county_fips),
  intensity_top %>% select(county_fips),
  by = "county_fips"
) %>%
  nrow()

tibble(overlap_top10 = overlap_n) %>% kable()
```

How to teach this section:

- If overlap is low, scale is materially changing ranking logic.
- If overlap is high, level and intensity are aligned for top counties.
- Neither result is "good" or "bad" by default; each tells you something about distribution.

### Teaching takeaway {.wb-takeaway}

This is where students usually move from "I ran code" to "I understand the empirical object."

## 10) Session 1 Close-Out Script

### Framing {.wb-framing}

If you want a clean closing script for this session, use this:

"Today we defined the target object, separated geography mistakes from valid measurement, and learned why scaling changes interpretation. We also built the habit of checking data structure before modeling."

### Checkpoint {.wb-checkpoint}

Before Session 2, make sure these can be explained without notes:

- What each term in the shift-share formula does
- Why the Laredo problem is a geography identification issue
- Why per-worker exposure is an intensity metric, not a replacement for levels
- Why early environment checks protect empirical integrity

<p class="small-note">Next session: rebuild exposure from components, run the NAICS problem scan, and turn maps into diagnostics rather than decoration.</p>
