---
title: "Tariff Exposure Teaching Series"
subtitle: "Session 3 - Regression Ladder, Uncertainty, and Optional Finance Controls"
author: "Ian Helfrich"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
    number_sections: false
    df_print: paged
    css: workbook_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

<div class="session-banner">
<b>Session focus:</b> Run a model ladder the right way: start simple, add structure, and report uncertainty honestly.
</div>

## 1) Environment and Packages

```{r env-check}
if (file.exists("retaliatory-tariff.Rproj")) {
  project_root <- "."
} else if (file.exists("../retaliatory-tariff.Rproj")) {
  project_root <- ".."
} else {
  stop("Could not find retaliatory-tariff.Rproj. Open this file inside the project folder.")
}

setwd(project_root)
knitr::opts_knit$set(root.dir = normalizePath(getwd()))
cat("Working directory:", normalizePath(getwd()), "\n")
```

```{r libraries}
needed <- c("readr", "dplyr", "stringr", "knitr", "fixest", "modelsummary")
missing_pkgs <- needed[!vapply(needed, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))]
if (length(missing_pkgs) > 0) {
  stop("Install missing packages first:\n", paste0("install.packages('", missing_pkgs, "')", collapse = "\n"))
}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(knitr)
  library(fixest)
  library(modelsummary)
})
```

## 2) Model Structure Refresher

<div class="math-legend">
  <span class="pill c-outcome">Green = dependent variable</span>
  <span class="pill c-share">Blue = political/context predictors</span>
  <span class="pill c-structure">Purple = fixed effects / clustering structure</span>
</div>

\[
\color{DarkGreen}{\log(1+\text{Exposure}_{ct})}
=
\beta_1 \color{MidnightBlue}{\text{GOP Share}_{c}}
+
\beta_2 \color{MidnightBlue}{\text{Swingness}_{c}}
+
\beta_3 \color{MidnightBlue}{\text{Tariffed Emp Share}_{ct}}
+
\color{DarkSlateBlue}{\text{state fixed effects}}
+
\varepsilon_{ct}
\]

### Framing {.wb-framing}

The point of the ladder is comparison across assumptions, not fishing for one model that looks best.

## 3) Load Data and Build Regression Frame

```{r load-panel}
panel <- read_csv(
  "data/public/derived/county_tariff_exposure_panel_with_votes_2017_2023.csv",
  show_col_types = FALSE
) %>%
  mutate(
    county_fips = str_pad(as.character(county_fips), 5, pad = "0"),
    year = as.integer(year),
    state_name = as.character(state_name),
    y = log1p(county_tariff_exposure_mean)
  )

reg_df <- panel %>%
  filter(
    is.finite(y),
    !is.na(gop_share_20),
    !is.na(swingness_20),
    !is.na(tariffed_emp_share)
  )

reg_df %>%
  summarise(
    rows = n(),
    counties = n_distinct(county_fips),
    years = n_distinct(year),
    states_with_labels = n_distinct(state_name[!is.na(state_name) & state_name != ""])
  ) %>%
  kable()
```

## 4) Regression Ladder

```{r regression-ladder}
m1 <- feols(
  y ~ gop_share_20 + swingness_20 + tariffed_emp_share,
  data = reg_df,
  vcov = "hetero"
)

m2 <- feols(
  y ~ gop_share_20 + swingness_20 + tariffed_emp_share | state_name,
  data = reg_df %>% filter(!is.na(state_name), state_name != ""),
  vcov = "hetero"
)

m3 <- feols(
  y ~ gop_share_20 + swingness_20 + tariffed_emp_share | state_name,
  data = reg_df %>% filter(!is.na(state_name), state_name != ""),
  vcov = ~state_name
)

modelsummary(
  list(
    "M1 HC1" = m1,
    "M2 + State FE (HC1)" = m2,
    "M3 + State FE (Clustered)" = m3
  ),
  stars = FALSE,
  fmt = 3,
  output = "markdown"
)
```

### Teaching takeaway {.wb-takeaway}

When confidence intervals widen under clustered SE, that is usually a realism correction, not a failure.

### Checkpoint {.wb-checkpoint}

Across M1 to M3, did signs change, magnitudes change, or uncertainty change most?

## 5) Optional Layer: Campaign Finance Controls

### Framing {.wb-framing}

This is an extension layer. It should pressure-test the main relationship, not replace the research question.

```{r finance-extension-check}
finance_panel_path <- "data/public/derived/county_tariff_panel_with_campaign_finance.csv"

if (!file.exists(finance_panel_path)) {
  cat(
    "Campaign-finance extension data is not present yet.\n\n",
    "To build it:\n",
    "1) source('code/add_campaign_finance_county_measures.R')\n",
    "2) Set FEC_API_KEY, or provide cached ZIP files in data/public/raw/\n",
    "3) Re-knit this file\n"
  )
} else {
  finance_panel <- read_csv(finance_panel_path, show_col_types = FALSE) %>%
    mutate(
      county_fips = str_pad(as.character(county_fips), 5, pad = "0"),
      year = as.integer(year)
    )

  needed <- c(
    "county_fips", "year", "county_tariff_exposure_mean", "gop_share_20",
    "swingness_20", "tariffed_emp_share", "state_name", "donations_total", "spending_total"
  )
  missing_needed <- setdiff(needed, names(finance_panel))

  if (length(missing_needed) > 0) {
    cat(
      "Finance panel is present, but required columns are missing:\n",
      paste0(" - ", missing_needed, collapse = "\n"),
      "\nRebuild with code/add_campaign_finance_county_measures.R\n"
    )
  } else {
    latest_finance_year <- max(finance_panel$year, na.rm = TRUE)
    fin_latest <- finance_panel %>% filter(year == latest_finance_year)

    fin_diag <- fin_latest %>%
      summarise(
        year = latest_finance_year,
        counties = n_distinct(county_fips),
        counties_with_donations = sum(donations_total > 0, na.rm = TRUE),
        counties_with_spending = sum(spending_total > 0, na.rm = TRUE),
        median_donations = median(donations_total, na.rm = TRUE),
        median_spending = median(spending_total, na.rm = TRUE)
      )

    kable(fin_diag, digits = 2, caption = "Campaign-finance coverage (latest year)")

    reg_fin <- fin_latest %>%
      mutate(
        y = log1p(county_tariff_exposure_mean),
        ln_donations = log1p(donations_total),
        ln_spending = log1p(spending_total)
      ) %>%
      filter(
        is.finite(y),
        is.finite(ln_donations),
        is.finite(ln_spending),
        !is.na(gop_share_20),
        !is.na(swingness_20),
        !is.na(tariffed_emp_share),
        !is.na(state_name),
        state_name != ""
      )

    m_base <- feols(
      y ~ gop_share_20 + swingness_20 + tariffed_emp_share | state_name,
      data = reg_fin,
      vcov = ~state_name
    )

    m_fin <- feols(
      y ~ gop_share_20 + swingness_20 + tariffed_emp_share + ln_donations + ln_spending | state_name,
      data = reg_fin,
      vcov = ~state_name
    )

    modelsummary(
      list(
        "Baseline FE + clustered" = m_base,
        "+ Donations + spending controls" = m_fin
      ),
      stars = FALSE,
      fmt = 3,
      output = "markdown"
    )
  }
}
```

## 6) Session Wrap and Writing Discipline

### Framing {.wb-framing}

Use this sequence every time you write findings:

- What changed in the specification
- What stayed stable in sign and rough size
- Where uncertainty widened
- What can be claimed as association
- What cannot be claimed as causal effect

### Checkpoint {.wb-checkpoint}

Draft a 4-6 sentence summary using M3 as the anchor model and M1/M2 as robustness context.

<p class="small-note">After this session, the next step is to package selected figures/tables into the client-facing narrative deck.</p>
